---
title: "Project 2"
author: "Gabriel Fraguere"
format: pdf
editor: source
---

## Section 1: Data Description

```{r}
#| label: load packages
#| warning: false
library(pokemon)

# Section 1
library(tidyverse)
library(tidymodels)

# Section 3 
library(tidyclust)

# Section 4 
#library(kknn)

```

```{r}
#| label: import data
library(pokemon) # This loads the `pokemon` dataset 
head(pokemon)
```
The data set I decided to choose was a data set containing information on 949 different Pokemon, across 7 different generations, while also including different forms of the same Pokemon, as well as "mega-evolutions" and other forms. This data set has exactly 22 variables. In this data set, each observation corresponds to a different Pokemon, which includes information regarding their stats. For example, the stats provided include the height, weight, base experience, type(s), health points, as well as their attack, defense, special attack, and special defense ratings, as well as other forms of data. The data was collected by Internet users "williamorim" and "marcusmunch" on GitHub, with the information being collected from the `Bulbagarden` Wikipedia page for Pokemon. 

```{r}
#| label: splitting the Pokemon dataset into a training and testing set 
set.seed(617, kind = "Mersenne-Twister")
poke.split <- initial_split(pokemon, prop = 0.75) # We are splitting with a proportion of 75% of the original data in the training set, and the other 25% set aside for the testing set. 
poke.train <- training(poke.split) # This creates our training set. 
poke.test <- testing(poke.split) # This creates our testing set. 

```

```{r}
#| label: checking the size of the training and testing sets 
nrow(poke.train) # This is how many observations are in our training set. 
nrow(poke.test) # This is how many observations are in our testing set. 

```

I decided to split the data set by a proportion of .75; This means that 75% the data set becomes our training set, which we'll use for creating and tuning models, and the remaining 25% was set aside for our testing set. Given that we have almost 1000 Pokemon in our data set, which range from regular, typical Pokemon, to stronger types and legendaries, I wanted a large enough number in our testing set to experiment with. We can also see that we have 710 observations in our training set, and 239 observations in the testing set. 

## Section 2: Data Wrangling and Exploratory Data Analysis

I would not like to be graded on the full `Data Wrangling and Feature Engineering`, but just the `Additional Exploratory Analysis` part.

Since our data set includes Mega Evolutions, Alola variants, etc, I want to do a little bit of data wrangling, and "clean" up our data set by removing these observations, given that these observations are simply "copies" of Pokemon already present in the data set. They share the same stats across the data set, with only a difference in their name. This is because in Pokemon, there exists other variants of the same Pokemon, such as `Pikachu` versus `alola-Pikachu`, with the latter just being from a different region. And so, we'll remove the Pokemon with dashes in their name, ("-"), and keep the rest.

```{r}
#| label: update data set 
pokemon.fix <- pokemon |>
  filter(!grepl("-", pokemon)) # The command `filter` filters out observations by checking if `pokemon$pokemon` contains a "-" in the name. We want to remove these Pokemon since they're all just other variants of already-existing Pokemon. 
nrow(pokemon.fix) # This shows us the new number of observations in the data set after data wrangling.
```

After removing the data, we now have 768 observations in our data set. From this, we'll also need to update our testing and training sets, since they are now affected by the data wrangling. I also want to stratify based on `base_experience`, given that stronger and legendary Pokemon have a higher value for this stat. I want to avoid the case where say, only legendaries/strong Pokemon are allotted into our testing set, which would affect our model. Stratification allows for similar proportions of `base_experience` in both `poke.train` and in `poke.test`.

```{r}
#| label: splitting the Pokemon dataset into a training and testing set

# Since `base_experience` is numeric and ranges from 0-605 in our data set, we have to break up the range of `base_experience` into different bins, so we can stratify later. 

pokemon.fix <- pokemon.fix |> 
  mutate(
    Exp_Lvl = cut(base_experience,
                  breaks = c(0, 100, 200, 300, Inf),
                  labels = c("Low", "Medium", "High", "Very High")
                  )
  )

set.seed(617, kind = "Mersenne-Twister")

poke.split <- initial_split(pokemon.fix, prop = 0.75, strata = Exp_Lvl) 
# We are still splitting with a proportion of 75% of the original data in the training set, and the other 25% set aside for the testing set; we are now also stratifying by on `Exp_Lvl`, which breaks up `base_experience` into four bins. 

poke.train <- training(poke.split) # We'll create our new testing and testing sets. 
poke.test <- testing(poke.split)

nrow(poke.train) # We'll take a look at the new size of our training and testing set. 
nrow(poke.test)  

```
We now have 575 observations in our training set, and 193 observations in our testing set. 

```{r}
#| label: look at data set 
glimpse(pokemon.fix)

```
I now want to explore if its possible to predict a Pokemon's `base_experience` stat based on it's other stats, (`attack`, `defense`, `speed`, etc); essentially, I want to determine if's possible to know the exact "strength" or "power" of a Pokemon off its stats alone. We should expect that with stronger Pokemon, they would show higher stats of `attack` and/or `speed`. I'll also be including `type_1`, to see if a Pokemon's type also has any influence on predicting `base_experience`. 

We'll start by looking at the distribution and five-number summary of `base_experience` as well as `hp`, or health points. By looking at the distribution of `base_experience`, we'll have a better idea of what to expect when trying to predict this response. I also chose `hp` as another variable/a predictor to look at, since I would also expect stronger Pokemon to have more health points. So, I would assume that their distribution relatively follows the same trend. 

```{r}
#| label: look at five number summary of base experience and hp

# While we can easily obtain a five-number summary by using summary(), I decided to manually create it here, since I wanted to include the standard deviation of each variable. 

poke.train |> 
  summarize(
    mean = mean(base_experience), # Mean 
    sd = sd(base_experience), # Standard deviation 
    min = min(base_experience), # Minimum
    q1 = quantile(base_experience, 0.25), # Q1 
    median = median(base_experience), # Median
    q3 = quantile(base_experience, 0.75), # Q3
    max = max(base_experience) # Maximum
  )

poke.train |> 
  summarize(
    mean = mean(hp), # Mean 
    sd = sd(hp), # Standard deviation 
    min = min(hp), # Minimum
    q1 = quantile(hp, 0.25), # Q1 
    median = median(hp), # Median
    q3 = quantile(hp, 0.75), # Q3
    max = max(hp) # Maximum
  )

```

From our results, we can see that `base_experience` has a mean of approximately 139.31, with a standard deviation of 73.03. It also appears that we do have an outlier at 608, given that that our third quartile is 175 ($608 > 339.25= 175 + (1.5)(109.5)$). Here, given that we have a large outlier relative to Q3, I would assume the distribution of `base_experience` to be bell-shaped, slightly right-skewed, with the data centered around the 130-180 range. 

For `hp`, we have a mean of 69.2, with a standard deviation of 27.6. Since our IQR is smaller than for `base_experience`, the spread of the data is more concentrated around the middle 50% of the data. We also have an outlier of 255, ($255 > 125 = 80 + (1.5)(30)$), which also implies right skewness. I would assume the distribution here to also be bell-shaped, not be as right-skewed, but not entirely symmetric given that our median and mean are not equal. 

For a better picture of the distribution, we'll plot both variables. 

```{r}
#| label: plotting base experience 
ggplot(
  data = poke.train, 
  mapping = aes( # This argument maps a specific aspect of the plot (x) to a particular variable in the data set (base_experience)
    x = base_experience 
  )
) + 
  geom_histogram(
    center = 125, # This argument refers to the center of one bin in the histogram. 
    binwidth = 10, # This argument refers to the width of each bin, which helps with reading the plot.
    fill = "olivedrab2" # This argument simply colors the bins. 
  ) + 
  labs( # This command creates labels for the title, x-axis, and y-axis. 
    x = "Base Experience",
    y = "Frequency of Pokemon with that base experience",
    title = "Pokemon and their base experience (according to Bulbapedia)"
  )

```
From here, we can see that the distribution of `base_experience` in our training set is moderately right skewed. The distribution peaks at around ~75-85, with roughly 160 Pokemon within that range. As we increase `base_experience`, we can see that there are less and less Pokemon with those levels. There also appears to be a second peak in the data, since there is also a good amount of data concentrated around the 150-180 range for `base_experience`; approximately 100 Pokemon in our training set are within this range for `base_experience`. This plot suggests that for our training set, there are two major concentrations of data, where these Pokemon hover around the same values for `base_experience`. Generally, I was mostly correct in my prediction for the distribution, but I didn't expect there to be two peaks. 

We'll continue and plot `hp`. 

```{r}
#| label: plotting hp  
ggplot(
  data = poke.train, 
  mapping = aes( 
    x = hp, 
  )
) + 
  geom_histogram(
    center = 125, 
    binwidth = 10, 
    fill = "firebrick4"
  ) + 
  labs(
    x = "HP",
    y = "Frequency of Pokemon with that HP",
    title = "Pokemon and their HP (according to Bulbapedia)"
  )

```
In comparison to the histogram for `base_experience`, the distribution of `hp` is much more unimodal, and almost symmetrical; there is some very light right-skewness, given that we have an outlier. The peak of the distribution is also much wider than I expected, which suggests that a considerable portion of Pokemon hover within this range for `hp`. To be more specific, it appears that at most Pokemon have `hp` levels concentrated within the range of 40-75, with a majority of the rest of the data just sitting outside this range, (mostly above, say ~75-125). I would say that I was pretty accurate in my prediction of this distribution: we have that it's approximately bell-shaped and almost symmetric, with some slight right skewness, and its centered about 40-75 for `hp`. 

Before we compare `base_experience` and `hp`, I want to also look at the distribution of `type_1`, which refers to the types of Pokemon within our data set, or specifically just how many of each type there are. From my understanding of Pokemon, I would assume that the types are even spread out among the observations, with less Pokemon for the later types, like Fairy, Steel, etc. 

```{r}
#| label: look at distribution and summary of type 1
poke.train |> 
  group_by(type_1) |>
  summarize(
    Freq = n() # This code will give us the number of observations/Pokemon with these types. 
  ) 

ggplot( # This will plot the different types of Pokemon and their cardinality. 
  data = poke.train,
  mapping = aes(
    x = type_1,
    fill = type_1
  )
) + 
  geom_bar() + 
  theme( # This command just helps with legibility when it comes to reading. 
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
      )
    ) + 
  labs(
    x = "Types",
    y = "Frequency of Pokemon with that (primary) type",
    title = "Pokemon and their (primary) type (according to Bulbapedia)"
  )

```
We can see that `bug`and `grass` types are the most prevalent within our training set, with there being 53 and 54 observations respectively. We have the least amount of `flying` types, with there being only 1, and the next smallest are `dragon`, `fairy`, and `dark`, respectively.  

From here, we'll plot a scatterplot of both variables (`base_experience` and `hp`), to see if there are any patterns/trends noticeable within their relationship. As I stated previously, I'd assume that "stronger" Pokemon, with a higher stat for `base_experience` would be correlated with a a high stat for `hp`; they would be a positive association between `base_experience` and `hp`. 

```{r}
#| label: plotting base experience and hp 
ggplot(
  data = poke.train, 
  mapping = aes(
    x = base_experience, 
    y = hp 
  )
) + 
  geom_point(
    alpha = 0.3,
    color = "darkgoldenrod2"
  ) + 
  geom_smooth(
    method = "loess", # This plots a LOESS curve to show the local trend for base_experience and hp
    se = F, # We don't really care for the confidence interval, so we'll leave it FALSE
    color = "purple4"
  ) + 
  labs(
    x = "Base Experience",
    y = "HP",
    title = "Relationship `base_experience` and `hp`"
  )

```
From our plot, we can see that the spread of the observations appears weak-moderately positively correlated, between `base_experience` and `hp`. There is also a lot of spread among the Pokemon, with outliers of `base_experience` AND `hp`, which appears to be the `blissey` and `chansey`. And so, there's weak evidence to suggest a positive association between `base_experience` and `hp`, with most of the Pokemon's stats varying heavily. 

I also want to explore the relationship of `base_experience` among the various types of Pokemon. Given that the sample sizes of types like `bug` and `normal` are our largest, their distributions would provide the best representation of the sample median for `base_experience` in our training set, with the rest of the types, like `flying`, `dragon`, and `fairy` having higher average `base_experience`. We'll plot a boxplot of each type, with their `base_experience` on the x-axis. 

```{r}
#| label: plotting base experience among the types of Pokemon 

ggplot(
  data = poke.train,
  mapping = aes(
    x = type_1, # Categorical must be on x-axis
    y = base_experience,
    fill = type_1 # colors the boxplots  
  )
)  + 
  geom_boxplot() + 
  geom_point(
    alpha = 0.1,
    position = position_jitter()
  ) + 
  scale_fill_manual( # chooses color for boxplots
    values = c("olivedrab","gray53", "tan3", "khaki1", "hotpink3", "beige", "orangered", "lightskyblue2", "lightcyan4", "palegreen4", "navajowhite4", "paleturquoise1", "seashell4", "aquamarine", "purple4", "gray80", "honeydew2", "dodgerblue")
  ) + 
  coord_flip() + 
  labs(
    x = "Base Experience",
    y = "Types of Pokemon",
    title = "Base Experience for each Pokemon type"
  )

```
From looking at the plot above, the first thing that stood out to me was that `dark`, `steel`, and surprisingly `fire` types all have the largest median `base_experience`, with `dragon` type having the largest IQR of all the types, ranging from 75 to upwards of 275. Compared to all of the types, `bug` and `normal` actually have a moderately sized boxes, which implies that there is much more spread among the middle 50% of these types. However, there are outliers present in the `normal` type which exceed an `base_experience` stat of 300, as well as outliers for the `dark` type which exceed a `base_experience` stat of 300. I would say that I was mostly accurate in my assumptions, seeing that the smaller sized types (`dark`, `steel`, `fire` surprisingly), have larger averages/concentrations of `base_experience` among the rest of the types. 

From my exploration of the data, I would argue that there isn't enough evidence to support a strong, positive association between `hp` and `base_experience`; but the two variables still share a weak-moderate positive relationship, with a few positive outliers. Initially I had expected the two to be very strongly, positively correlated, assuming that "stronger" Pokemon would have more health, but I now question whether `base_experience` is a strong quantifier or not for "strength". Additionally, while I did expect the "newer" types to have higher stats of `base_experience`, since they are "rarer" and are more likely to be "stronger", (by my basic understanding of Pokemon), I didn't expect `fire` type to also show a higher median for `base_experience`, seeing as they didn't have the largest sample size. 

## Section 3: Unsupervised Learning

I decided to implement k-means clustering on our `poke.train` data set, to see if there exists any noticeable "clusters" or groups of observations with similar characteristics. This technique seeks to assign each observation to a specific group assignment, that should share similarities (assign each Pokemon to a group based on similarities in the data). 

This algorithm works by choosing some number of of cluster assignments (the value of $k$), computing a vector of averages for each feature, for each cluster, then assigning each observation/Pokemon to cluster with closest stats to the vector (known as a centroid). 

To determine the number of clusters, I used a for-loop to test different values for the numbers of clusters, from 1 to 8. I'll plot the corresponding elbow plot to determine the optimal value. 

```{r}
#| label: determine how many clusters 

set.seed(617)

# Step 1: Scale numeric predictors 
poke.scaled <- poke.train |>
  select(hp, attack, defense, special_attack, special_defense, speed) |> 
  scale()

# Step 2: Create a sequence of values to test, as well as a placeholder object for the total within cluster sums of squares. 
k_vals <- 1:8  
tot_withinss_vals <- numeric(8)

# Step 3: Use for loop to test and collect tot_within values. 
for (i in seq_along(k_vals)) {
  k <- k_vals[i]
  poke.km.fit <- kmeans(
    poke.scaled,
    centers = k,
    nstart = 20
  )
tot_withinss_vals[i] <- poke.km.fit$tot.withinss
}

# Step 4: Compile the values we tested for the number of clusters and their corresponding total within cluster sum of squares. 
poke.elbow <- data.frame(
  k = k_vals,
  tot_withinss = tot_withinss_vals
)

# Step 5: Plot the elbow curve based on the values we tested and their tot_withinss values to determine the best value for k.
ggplot(
  poke.elbow, 
  aes(
    x = k,
    y = tot_withinss
    )
  ) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Plot for Choosing Number of Clusters",
    x = "Number of Clusters",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal()
```
From our plot, I decided to choose 3 clusters, since at this point, the amount that we're decreasing our total within sum of squares by is decreasing, as in that increased clusters wouldn't provide as substantial of a benefit. I would say that we could use 3 or 4 clusters for our training set, but I just decided to go with 3.

I also want to touch on why I excluded `type_1` from when implementing k-means clustering. This is due to the fact that k-means clustering relies on Euclidean distance, which is best suited for continuous, numeric variables. If `type_1` was included as a binary variable, our elbow plot may be distorted. 

```{r}
#| label: set up kmeans model and workflow 
set.seed(617)

# We'll create a new k-means clustering model
poke.km.model <- k_means(num_clusters = 3) |> 
  set_engine("stats") |>
  set_args(nstart = 20) # This just means we'll run from 20 random initial assignments 

# We'll also create a new recipe containing
poke.km.recipe <- recipe(~ hp + attack + defense + special_attack + special_defense + speed, 
                           data = poke.train) |>
  step_dummy(all_nominal_predictors()) |> # This converts categorical variables to dummy variables 
  step_normalize(all_numeric_predictors()) # This standardizes our variables

poke.km.wflow <- workflow () |> # We'll create the new workflow containing our new clustering
  add_model(poke.km.model) |> 
  add_recipe(poke.km.recipe)
```

```{r}
#| label: fit workflow to training data and look at the clusters 
poke.km.fit <- poke.km.wflow |> # We'll fit the workflow to our training set
  fit(data = poke.train)

poke.km.aug <- poke.km.fit |> # We'll augment our training set by adding the predicted clusters for each Pokemon (observation)  
  augment(new_data = poke.train)
poke.km.aug

poke.km.clusters <- poke.km.aug |> # This line creates and displays the clusters. 
  group_by(.pred_cluster) |>
  summarise(
    hp = mean(hp),
    attack = mean(attack),
    defense = mean(defense),
    special_attack = mean(special_attack),
    special_defense = mean(special_defense),
    speed = mean(speed),
    n = n()
  )
poke.km.clusters

```

From our summaries, we can see that each observation/Pokemon in our training set has a predicted cluster assigned to it; for example, Charizard and Venusauar are all assigned to Cluster 1. 

Looking at the clusters themselves, we have that Cluster 1 has the highest averages for `attack` (92.00), `special_attack` (90.59), and `speed` (92.70). This cluster is also the second largest, beating out cluster 2 by 45 Pokemon. This cluster likely contains the offensively-strong Pokemon, or the "heavy-hitters;" these Pokemon, while not strong `hp`-wise or defensively, their attacks are the strongest and the quickest.  

On the other hand, Cluster 3 has the lowest averages for ALL stats, with the exception of sample size; of which, Cluster 3 is the largest, with 238 Pokemon in our training set assigned to it. Their averages are all in the 50s, implying that this group refers to the much weaker and basic Pokemon within our training set; the fodder that a trainer would farm after randomly encountering it in the wild.

There is also Cluster 2, who has highs in `hp` (87.82), `defense` (97.99), and `special_defense` (85.06). This is also the smallest cluster, with only 146 Pokemon assigned to this group. This cluster refers to the defensively-strong Pokemon; they are much more resistant to attack and have much more sustain than the other Pokemon in our training set. 

```{r}
#| label: plot clusters
# Here, we'll create a scatter plot of all the observations, based on two variables (speed and defense), colored based on their cluster assignment. I chose these two variables since they felt like strong enough features to help distinguish between the clusters. 

ggplot(
  poke.km.aug,
  aes(
    x = speed, 
    y = defense, 
    color = .pred_cluster)
) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Pokémon Clusters Based on Speed and Defense",
    x = "Speed",
    y = "Defense",
    color = "Cluster Assignment"
  ) +
  theme_minimal()

```
The scatterplot shown clearly shows the three different clusters. Cluster 1 (Offense-Strong Pokemon) are colored red; their data points are relatively higher than the rest of the Pokemon for `speed` (75-125), but the majority of Cluster 1's data points fall between the 50-100 range for `defense`. For Cluster 2 (Defense-Strong Pokemon), their data points lie lower for `speed` (25-75), but higher for `defense` (75-150). Touching on Cluster 3, their data points/Pokemon all lie towards the lower-left of the graph, which indicate lower values for `defense` and `speed`, supporting the interpretation that this group contains the "weaker" Pokemon.  

## Section 4: Supervised Learning

For the first model, I decided to fit a k-nearest neighbors model on the training data, trying to predict `base_experience` based on a tuning parameter for the value of $k$. Our data set doesn't necessarily have a natural response variable, so I chose `base_experience` for the sake of attempting to predict a Pokemon's base "strength" based on their stats alone. 

The value $k$ represents the number of neighbors 

Having too many neighbors -> (large k) 
Not enough neighbors -> (small k)

We begin by first performing 

```{r}
#| label: k-fold cross validation set up 
set.seed(617)
poke.cv <- vfold_cv(
  poke.train, # training set
  v = 5 # five folds
)

```

We'll then

```{r}
#| label: create and add to workflow 
poke.knn.model <- nearest_neighbor(
  mode = "regression",
  engine = "kknn",
  neighbors = tune(), # this is the parameter we're tuning, since we want to determine an optimal value for "k" 
  dist_power = 1
)

poke.knn.recipe <- recipe(
  base_experience ~ type_1 + hp + attack + defense + special_attack + special_defense + speed, # This 
  data = poke.train 
) |>
  step_novel() |>
  step_dummy(all_nominal_predictors()) |> # This 
  step_normalize(all_numeric_predictors()) # This 

poke.knn.wflow <- workflow() |> 
  add_model(poke.knn.model) |> 
  add_recipe(poke.knn.recipe)


```

Next we'll 

```{r}
#| label: set up tuning grid for neighbors and dist power 
poke.tune.grid <- expand.grid(
  neighbors = seq(1, 100, by = 25)
)

poke.knn.tuned <- tune_grid(
  poke.knn.wflow,
  resamples = poke.cv,
  grid = poke.tune.grid, 
  metrics = metric_set(rmse)
)

poke.knn.bestk <- select_by_one_std_err(
  poke.knn.tuned, 
  metric = "rmse",
  neighbors)
poke.knn.bestk

```
From our results, we can see that 

```{r}
#| label: plot tuned model and look at average accuracy / sd of accuracy
autoplot(poke.knn.tuned) 
poke.knn.results <- collect_metrics(poke.knn.tuned)|>
  mutate(sd = std_err * sqrt(5)) # This 
poke.knn.results

```
As the plot show, at ... .... ..., and so selecting $k=5$ gives us ____ RMSE, impling that 

Our output also shows that a value of ___ has a mean RMSE of, with a standard deviation 

```{r}
#| label: finalize model and workflow
poke.knn.f.wflow <- poke.knn.wflow |> 
  finalize_workflow(
    poke.knn.bestk
  )

poke.knn.fit <- fit(poke.knn.f.wflow,
                     data = poke.train)
poke.knn.fit
```

Following this, I also want to fit a linear regression model on the data, to see there exists any 

```{r}
#| label: create and add model to workflow 
poke.lm.model <- linear_reg(
  mode = "regression", 
  engine = "lm"
)

poke.lm.wflow <- workflow() |> 
  add_model(poke.lm.model) |>
  add_recipe(poke.knn.recipe) # We're using the same recipe 

poke.lm.cv <- fit_resamples(
  poke.lm.wflow,
  resamples = poke.cv,
  metrics = metric_set(rmse)
)
  
poke.lm.results <- collect_metrics(poke.lm.cv)|>
  mutate(sd = std_err * sqrt(5)) # This

```

```{r}
#| label: compare accuracy and sd of accuracy for both models 
poke.lm.results
poke.knn.results 
```

From our results, we can see that 